-----------------------------------------------
title: "Reproducbile Research Project 1"
author: "Samantha Foster"
date: "Friday, June 12, 2015"
-----------------------------------------------

first, load the required libraries

```{r}
library(knitr)
library(dplyr)
library(lattice)
```


###Question 1: Loading and preprocessing the data

Unzip the activity.zip file and read the .csv file into a variable called data

```{r}
setwd("C:/Users/Sam/Documents/Coursera/ReproducibleResearch/Project1/RepData_PeerAssessment1")
unzip("activity.zip")
data<-read.csv("activity.csv",header=TRUE,sep=",")
```

Convert the interval variable into a factor for use in later analysis.  This is the only processing/transforming needed right now.

```{r}
data$interval <- as.factor(data$interval)
```

###Question 2: What is the mean total number of steps taken per day?

To calculate the total number of steps taken per day, group the data by day using 
the group_by function in dplyr, then use the summary function to calculate the sum 
of steps for each day

```{r}
dayGroup <- group_by(data,date)
daySteps <- summarize(dayGroup, stepsPerDay = sum(steps, na.rm =TRUE))
```

Make a historgram of the total number of steps per day
```{r}
hist(daySteps$stepsPerDay, breaks = 20, main = "Histogram of Steps Taken per Day", xlab = "Steps Per Day")
```

Calculate and output the mean and median steps per day
```{r}
meanSteps <- round(mean(daySteps$stepsPerDay),0)
meanSteps
medianSteps <- median(daySteps$stepsPerDay)
medianSteps
```

The mean of the total number of steps per day is `r meanSteps`, and the median is `r medianSteps`.

###Question 3: What is the average daily activity pattern?

Group the data by interval using the group_by function in dplyr, then use summary to calculate the 
average number of steps taken for each interval, averaged across all days

```{r}
intervalGroup <- group_by(data,interval)
intervalSteps <- summarize(intervalGroup, stepsPerInterval = mean(steps, na.rm=TRUE))

```

Make a plot of the interval and average number of steps.  

```{r}
intervalSteps$interval <- as.numeric(levels(intervalSteps$interval))
plot(intervalSteps$interval, intervalSteps$stepsPerInterval, main = "Average Number of Steps per Interval",xlab="Inverval", ylab="Average Steps",type="l")
```

Calculate the interval that contains the maximum number of steps
```{r}
maxIndex <- match(max(intervalSteps$stepsPerInterval),intervalSteps$stepsPerInterval)
maxInterval <- intervalSteps$interval[maxIndex]
maxInterval
```

The interval with the maximum number of steps is `r maxInterval`.

###Question 4: Inputing missing values 

Calculate the total number of missing values in the steps column of the data
```{r}
totalNA <- length(which(is.na(data$steps)))
totalNA
```
The total number of missing values is `r totalNA`

If a value is missing, replace it with the mean value for the associated 5-minute interval (calculated in Question 3, and stored in intervalSteps variable.
Make a new dataset called data2, that has all NA values replaced.

```{r}
data2 <- data

for(i in 1:length(data$steps)){
  if(is.na(data$steps[i])){
    temp <- data$interval[i]
    index <- match(temp,intervalSteps$interval)
    steps <- intervalSteps$stepsPerInterval[index]
    data2$steps[i] <- steps
  }
}

```

Repeat question 2, but using data2


```{r}
dayGroup2 <- group_by(data2,date)
daySteps2 <- summarize(dayGroup2, stepsPerDay2 = sum(steps, na.rm =TRUE))
```

Make a historgram of the total number of steps per day
```{r}
hist(daySteps2$stepsPerDay2, breaks = 20, main = "Histogram of Steps Taken per Day", xlab = "Steps Per Day")
```

Calculate and output the mean and median steps per day
```{r}
meanStepsNA <- round(mean(daySteps2$stepsPerDay2),0)
meanStepsNA
medianStepsNA <- round(median(daySteps2$stepsPerDay2),0)
medianStepsNA
```

The mean of the total number of steps per day with NAs inputed is `r meanStepsNA`, and the median is `r medianStepsNA`.

Calculate the difference in mean and median steps from the first part of the assignment.

```{r}
meanDiff <- abs(meanSteps - meanStepsNA)
meanDiff
medianDiff <- abs(medianSteps - medianStepsNA)
medianDiff

```

The difference in mean steps per day between ingorning NA values, and adding them in is `r meanDiff`, and the median is `r medianDiff`. The impact of adding missing data is that the mean and median of daily number of steps are now equal. 

###Question 5: Are there differences in activity patterns between weekdays and weekends?

First, change the date column from a factor into a date
```{r}
data2$date <- as.Date(data2$date)
```

Add a new factor variable column to the data2 dataframe that indicates if the date is a weekday or weekend.  Use the mutate function from dplyr to do this

```{r}
data2<- mutate(data2, day = factor((weekdays(date)=="Saturday"| weekdays(date)=="Sunday"), labels = c("weekdays","weekend")))
```

Group the data by interval using the group_by function in dplyr, then use summary to calculate the 
average number of steps taken for each interval, averaged across all days.  This will be used for the plot

```{r}
intervalGroup3 <- group_by(data2,interval,day)
intervalSteps3 <- summarize(intervalGroup3, stepsPerInterval3 = mean(steps, na.rm=TRUE))
```

Make a pannel plot of the steps per interval for weekends and weekdays
```{r}

xyplot(stepsPerInterval3~interval | day, data = intervalSteps3, layout=c(1,2),ylab="Number of steps",type="l")
```
